---
title: "Introduction to the autocart package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to the autocart package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The autocart package is a version of the classification and regression tree algorithm, but adapted to explicitly consider measures of spatial autocorrelation inside the splitting itself. The autocart package will be of most use for ecological datasets that cover a global spatial process that can't be assumed to act the same at every local scale.

To get started, load the library into R.

```{r message = FALSE, warning = FALSE}
library(autocart)
library(sp)
```

## Example dataset preparation

```{r dataset}
# Load the example snow dataset provided by the autocart package
snow <- read.csv(system.file("extdata", "ut2017_snow.csv", package = "autocart"))
```

The provided snow dataset contains measures of ground snow load at a variety of sites located in Utah. The response value "yr50" contains the ground snow load value, and a variety of other predictor variables are provided at each of the locations.

```{r}
head(snow)
```

There are a couple NA values in this data set. If we pass in a data set with missing information, it will throw an error.. For this vignette, we will choose to remove all the rows that contain any sort of missing observations.

```{r}
snow <- na.omit(snow)

# Pick just a few important variables
snow <- snow[, c("yr50", "ELEVATION", "MWMT", "PPTWT", "LONGITUDE", "LATITUDE")]
```

To function properly, autocart needs a SpatialPointsDataFrame() (instead of the normal data.frame) passed in.

```{r}
snow <- SpatialPointsDataFrame(coords = snow[, c("LONGITUDE", "LATITUDE")],
                               data = snow)
```


Let's split the data into 85% training data and 15% test data. We will create a model with the training data and then try to predict the response value in the test data set.

```{r}
# Split the data into 85% training data and 15% test data
numtraining <- round(0.85 * nrow(snow))
training_index <- sample(c(rep(TRUE, numtraining), 
                           rep(FALSE, nrow(snow) - numtraining)))
snow_train <- snow[training_index, ]
snow_test <- snow[!training_index, ]
```

## Training a model

One crucial parameter we pass into autocart is the "alpha" parameter. Inside of the splitting function, we consider both a measure of reduction of variance, as well as a statistic of spatial autocorrelation. We can choose to weight each of the measures different. The alpha value that we pass into the autocart function says how much the splitting function will weight the statistic of spatial autocorrelation. If we set alpha to 1, then we will only consider autocorrelation in the splitting. If we set alpha to 0, then autocart will function the same as a normal regression tree. For this example, let's set alpha to be 0.30.

```{r}
alpha <- 0.30
```

Another parameter we can weight is "beta", which ranges from a scale of 0 to 1. This controls the shape of the regions that are formed. If beta is near 1, then the shapes will be very close together and compact. If beta is 0, then this shape will not be considered in the splitting. Let's pick beta to be 0.05.

```{r}
beta <- 0.05
```

If we want extra control over how our tree splits, we can also mess with the autocartControl() function. As an example of changing a deeper parameter, let's make it so that 70\% of the maximum distance between points is the maximum distance at which spatial influence between points is assumed.

```{r}
my_control <- autocartControl(spatialBandwidthProportion = 0.70)
```

Finally, we can create our model!

```{r}
snow_model <- autocart(yr50 ~ ELEVATION + MWMT + PPTWT, alpha = alpha, beta = beta, 
                       data = snow_train, control = my_control)
```

The autocart function returns an S3 object of type "autocart". We can use the predict() function to use the object to make predictions for the testing dataset.

```{r}
test_predictions <- predict(snow_model, snow_test) 
```

We can see how well we did by getting the root mean squared error.

```{r}
residuals <- snow_test$yr50 - test_predictions

# RMSE
sqrt(mean(residuals^2))
```
